---
title: "Solutions to the STL R-Ladies Workshop"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Source
These exercises are borrowed and adapted from the [R for Data Science book](https://r4ds.had.co.nz/) and Thomas Pedersen's workshop on GGPLOT2. The latter is available on [youtube](https://www.youtube.com/watch?v=0m4yywqNPVY) and a great resource if you want to explore further.   

# Rmd Document
This document contains all the code for exercises. The
document is an RMarkdown document which means that it can be compiled, along 
with the code chunks thus executing and capturing the output of the code within
the document. To read more about RMarkdown see the website for the package, as
well as the [Get Started](https://rmarkdown.rstudio.com/lesson-1.html) guide.

### Dependencies
List of required packages is below. If library(package name) throws an error, install the package first using `install.packages("package name")`. Go ahead and run the chunk below by pressing the green button on the right. This will load all the required libraries.   

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(sf)
library(maps)
```

#### Exercise

Modify the code below to make the points larger triangles and slightly transparent.
See `?geom_point` for more information on the point layer.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy))
```

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy), shape="triangle", size=4, alpha=0.5)
```

* * *

Using the `mpg` dataset draw a line chart, a boxplot, and a histogram

```{r}
ggplot(data=mpg)+
  geom_line(aes(x=displ, y=hwy))
ggplot(data=mpg)+
  geom_boxplot(aes(x=class, y=displ))
ggplot(data=mpg)+
  geom_histogram(aes(x=hwy))
```

### Stat

###Exercises
What does geom_col() do? How is it different from geom_bar()?

Look at the documentation for geom_bar using `?geom_bar`

* * *

We learnt that `geom_*()` and `stat_*()` are interchangeable. Can you look at `?geom_bar()` and figure out which
stat it uses as default. Modify the code below to use that stat directly instead

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class))
```


The description says "geom_bar() uses stat_count() by default". Using it directly below:

```{r}
ggplot(mpg) + 
  stat_count(aes(x = class))
```

* * *
Use `stat_summary()` to add a red dot at the mean `hwy` for each group

```{r}
ggplot(mpg) + 
  geom_jitter(aes(x = class, y = hwy), width = 0.2)
```
Hint: You will need to change the default geom of `stat_summary()`

```{r}
ggplot(mpg, aes(x=class, y=hwy)) + 
  geom_jitter(width = 0.2)+
  stat_summary(geom = "point", fun="mean", color="red")
```


* * *

In our proportion bar chart, we need to set group = 1. Why? In other words what is the problem with these two graphs?
```{r}
p1<- ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop)))

p2<-ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop)))

p1
p2
```

if group = 1 is not included, the proportions will be calculated within each group

```{r}
p1_new<- ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop), group=1))

p2_new<-ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = ..count.. / sum(..count..)))

p1_new
p2_new
```
***
What is the problem with this plot? How could we improve it?

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point( )
```

There is overplotting because there are multiple observations for each combination of `cty` and `hwy` values. 

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point(position="jitter" )
```

### Scales

#### Exercises
Use `RColorBrewer::display.brewer.all()` to see all the different palettes from
Color Brewer and pick your favourite. Modify the code below to use it

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(type = 'qual')
```

```{r}
data("mpg")
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(type = 'qual', palette = "Set1")
```
* * *

Modify the code below to create a bubble chart (scatterplot with size mapped to
a continuous variable) showing `cyl` with size. Make sure that only the present 
amount of cylinders (4, 5, 6, and 8) are present in the legend.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(type = 'qual')
```

Hint: The `breaks` argument in the scale is used to control which values are
present in the legend.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class, size=cyl)) + 
  scale_colour_brewer(type = 'qual') +
  scale_size(breaks = c(4, 5, 6, 8))
```
Explore the different types of size scales available in ggplot2. Is the default
the most appropriate here?

Default is mapping to the radius. But it is not intuitive. Let's try size mapping by area. 
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class, size=cyl)) + 
  scale_colour_brewer(type = 'qual') +
  scale_size_area(breaks = c(4, 5, 6, 8))
```
* * *

Modify the code below so that colour is no longer mapped to the discrete `class`
variable, but to the continuous `cty` variable. What happens to the guide (legend)?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class, size = cty))
```

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = cty, size = cty))
```
* * *

The type of guide can be controlled with the `guide` argument in the scale, or 
with the `guides()` function. Continuous colours have a gradient colour bar by 
default, but setting it to `legend` will turn it back to the standard look. What 
happens when multiple aesthetics are mapped to the same variable and uses the 
guide type?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = cty, size = cty))+
  guides(color="legend")
```

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = cty, size = cty))
```

### Facets

#### Exercises
One of the great things about facets is that they share the axes between the 
different panels. Sometimes this is undesirable though, and the behavior can
be changed with the `scales` argument. Experiment with the different possible
settings in the plot below:

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv)
```

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv, scales="free_y")
```
* * *

Usually the space occupied by each panel is equal. This can create problems when
different scales are used. Can you modify the code below so that the y scale differs 
between the panels in the plot. What happens?

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = manufacturer)) + 
  facet_grid(class ~ .)
```

Use the `space` argument in `facet_grid()` to change the plot above so each bar 
has the same width again.

```{r}
data("mpg")
ggplot(mpg) + 
  geom_bar(aes(y = manufacturer)) + 
  facet_grid(class ~ ., space = "free_y", scales = "free_y")
```

* * *
### Coordinates
The coordinate system is the fabric you draw your layers on in the end. The 
default `coord_cartesion` provides the standard rectangular x-y coordinate 
system. Changing the coordinate system can have dramatic effects
```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class)) + 
  coord_polar()
```

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class)) + 
  coord_polar(theta = 'y') + 
  expand_limits(y = 70)
```

You can zoom both on the scale...

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class)) + 
  scale_y_continuous(limits = c(0, 40))
```

and in the coord. You usually want the latter as it avoids changing the plotted
data

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class)) + 
  coord_cartesian(ylim = c(0, 40))
```

#### Exercises
In the same way as limits can be set in both the positional scale and the coord,
so can transformations, using `coord_trans()`. Modify the code below to apply a
log transformation to the y axis; first using `scale_y_continuous()`, 
and then using `coord_trans()`. Compare the results — how do they differ?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = hwy, y = displ))
```


```{r}
ggplot(mpg) + 
  geom_point(aes(x = hwy, y = displ))+scale_y_continuous(trans = "log")
```

```{r}
ggplot(mpg) + 
  geom_point(aes(x = hwy, y = displ))+
  coord_trans(y="log")
```
The break points have been calculated on the normal data here, so breaks are more sensible here. 
* * *

Coordinate systems are particularly important in cartography. While we will not 
spend a lot of time with it in this workshop, spatial plotting is well supported
in ggplot2 with `geom_sf()` and `coord_sf()` (which interfaces with the sf 
package). The code below produces a world map. Try changing the `crs` argument 
in `coord_sf()` to be `'+proj=robin'` (This means using the Robinson projection). 

```{r}
# Get the borders of all countries
world <- sf::st_as_sf(maps::map('world', plot = FALSE, fill = TRUE))
world <- sf::st_wrap_dateline(world, 
                              options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180"),
                              quiet = TRUE)
# Plot code
ggplot(world) + 
  geom_sf() + 
  coord_sf(crs = "+proj=moll")
```

```{r}
# Plot code
ggplot(world) + 
  geom_sf() + 
  coord_sf(crs = "+proj=robin")
```

### Theme

#### Exercises
Themes can be overwhelming, especially as you often try to optimise for beauty 
while you learn. To remove the last part of the equation, the exercise is to 
take the plot given below and make it as hideous(!) as possible using the theme
function. Go absolutely crazy, but take note of the effect as you change 
different settings. To get started look at the documentation `?theme` and pick ten (or more) arguments to modify. 

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = class, fill = drv)) + 
  facet_wrap(~year) + 
  labs(title = "Number of car models per class",
       caption = "source: http://fueleconomy.gov",
       x = 'Number of cars',
       y = NULL)
```






